\bf Causal Inference in Population Trends: Searching for Demographic Anomalies in Big Data

### 2021-03-29 MainDocument/MainDocument.tex
Updated.

@@

  \author{
        Mathew E. Hauer \thanks{Thanks y'all!} \\
    Department of Sociology, Florida State University\\
     and \\     Stephanie A. Bohon \\
    Department of Sociology, University of Tennessee - Knoxville\\
      }
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
The proliferation of big data, wider access to advanced computing
platforms, and the development of powerful statistical algorithms can
uncover hidden anomalies in social data, previously dismissed as noise.
Here, we combine causal inference techniques and abductive reasoning to
identify fertility and mortality anomalies on twenty years of complete
demographic data in the United States. We uncover real, ``hidden'' baby
booms/busts and mortality spikes/dips, distinguishable from regular
trend variations. We identify more than 22 and 156 fertility and
mortality anomalies, totaling more than 200k and 600k anomalous births
and deaths, respectively. Notable detectable mortality anomalies include
the September 11 2001 terrorist attack in New York, the emergence and
acceleration of the opioid epidemic in New Hampshire, and the excess
mortality due to Hurricane Maria in Puerto Rico. Notable fertility
anomalies include the ``missing births'' in Louisiana after Hurricane
Katrina and the reduction in fertility behavior after the September 2008
stock market crash in Connecticut, amongst others. Our combined causal
inference and abductive reasoning approach can be readily adapted to
find other, undiscovered social phenomena or to evaluate the efficacy of
important public policies.
\end{abstract}

@@

The search for causality is one of the primary motivations behind social
science \citep{smith2011commentary, pearl2009causal}. Causal Inference,
the identification of causal mechanisms, attempts to uncover the
underlying mechanisms that result in changes in a phenomenon and has a
long history as an approach to inquiry in the social sciences
\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
is accomplished via a randomized control trial (RCT) or expert knowledge
applied to a natural experiment \citep{salganik2019bit}.

Unfortunately, both RCTs and natural experiments have drawbacks. RCTs
are often prohibitively expensive for many scientists, precluding their
widespread adoption in the social sciences \citep{west2008ajph}. RCTs
are also sometimes impossible to deploy in the ``real world''. For
instance, an RCT to study country-level austerity measures would be
difficult, if not impossible, to implement. Natural experiments are a
cheaper alternative, but provide less certainty about causation, fewer
opportunities to replicate work, and require expert knowledge of the
natural experiment to be properly employed \citep{pearl2018book}.

Additionally, both RCTs and natural experiments follow traditional
hypothesis testing, the deductive scientific paradigm where a question
is first posed and scientists generate or find data to answer the
question. Such approaches, while extensively utilized and time-tested,
are likely to miss important phenomena that might go unnoticed or suffer
from extensive p-hacking
\citep{head2015extent, nuzzo2014scientific, ruggles2014big}. Abductive
reasoning, sometimes described as ``inferring cause from effect''
\citep{Crowder2017}, moves the scientific process from the inductive to
the deductive, and sometimes back and forth, to reach conclusions
\citep{bryant2014realm} The physical sciences use abductive reasoning
more fully compared to the social sciences. The discovery of Neptune in
1846 based on unexplained perturbations of the orbit of Uranus
\citep{popper2005logic}, the discovery of galactic migration of planets
\citep{gomes2005n}, and the current generation of gravitational wave
detectors such as the Laser Interferometer Gravitational-Wave
Observatory (LIGO) \citep{harry2010advanced} demonstrate the power of
the abductive approach in the physical sciences.

In a data-rich environment, abductive reasoning can be more easily
applied to first find effects and then infer their cause ex-post. The
modern proliferation of data, advances in high performance (``super'')
computing, and the development of powerful statistical algorithms
provide an environment ripe for wedding causal inference with abductive
reasoning \citep{van2016data, zikopoulos2011}, with the potential for
researchers to find hidden or understudied social phenomena
\citep{bohon2018demography}. While the availability of Big Data and high
performance computing allows novel exploration of data through causal
inference
\citep{bohon2018demography, rcausalimpact, shiffrin2016drawing},
relatively few studies utilize causal inference techniques in the study
of demographic phenomena. However, understanding social phenomena using
these advances reveals important insights into society
\citep{angrist1989lifetime, mas2009peers} and allows us to better
monitor population trends \citep{nobles2019, torche2015hidden}. The rich
demographic data available in the United States make the potential
revelation of interesting and important demographic phenomena not only
possible, but extremely plausible -- even if identification of the
phenomena occurs without identifying the underlying cause.

In this paper, we combine a modern statistical outlier detection
algorithm \citep{chen1993joint}, abductive reasoning, and large
demographic data sets on more than 20 years of mortality/fertility to
identify anomalous demographic behavior. In essence, we identify
\emph{effects without knowledge of the cause}. We ask one fundamental
question regarding demographic anomalies: What are the hidden baby
booms/busts and mortality spikes/dips in the United States over the last
twenty years? We do not necessarily know the causes of these anomalies
but identifying them allows scientists with more detailed knowledge of
local population dynamics, state-level policy making, or macro-economics
to explain these phenomena post-hoc. From explanations of phenomena that
may have previously gone unnoticed, social scientists may be able to
better forecast populations and provide policy solutions for impending
problems such as climate change.

@@

Scientific inquiry falls under one of three modes of logical reasoning:
deduction, induction, and abduction. Deductive reasoning starts with a
theory and makes a prediction of what the observations should be if that
theory is correct. The classic example of deductive reasoning
essentially goes (a) ``all men are mortal'', (b) ``Socrates is a man'',
(c) ``Therefore, Socrates is mortal.'' Deductive reasoning has a long
history in the sciences and most hypothesis testing in the social
sciences follows this logical reasoning. For example, John Snow used
deductive reasoning in to stop the 1854 London cholera outbreak. He
hypothesized that cholera was waterborne and deduced that a bad water
pump was responsible for the outbreak.

Inductive reasoning is the opposite of deductive reasoning. Rather than
starting with a theory and making a prediction, induction starts with an
observation and makes a theory based on the likeliest reason. For
example, the Hispanic Mortality Paradox first appeared in the scientific
literature in the mid-1980s \citep{markides1986health} as an observation
that hispanics in the United States had, paradoxically, better health
outcomes than expected given their socioeconomic status. Based on this
observation, numerous theories arose to explain this paradox.

Both deductive and inductive reasoning are well established in science
but are not without their shortcomings. In both approaches, it is
impossible, or at least imporbable, for the premise to be true and the
conclusion false. Furthermore, if a hypothesis is rejected under a
deductive or inductive approach, there is no alternative explanation --
there is either the presence or absence of support.

@@

Abductive reasoning and exploratory data analysis are best employed in a
data rich environment -- the very data environment social scientists and
demographers find themselves in the present moment. Additionally, in a
data rich environment, a hypothesis is a hidden liability
\citep{yanai_hypothesis_2020}because it focuses our attention on only
that which we seek\footnote{Yanai and Lercher
  \citeyearpar{yanai_hypothesis_2020} ask if focusing on a specific
  hypothesis prevents new discovery? They created a simple dataset that
  if plotted would reveal an image of a gorilla waving. Students
  instructed to find support for specific hypotheses were less likely to
  discover the gorilla compared to those who were simply asked ``What do
  you conclude from the dataset?''}. Focusing on specific hypotheses
often prevents us from finding other interesting phenomena. Afterall, it
was the publication of the hispanic paradox itself that has generated
more than thirty years of devoted scholarship. When paired with causal
inference techniques, abductive reasoning provides an empirical platform
for the discovery of potentially interesting social phenomena; it
provides an important tool for sifting large amounts of data to find
interesting results in the absence of hypothesis testing.

@@

To determine the false-positive (Type I) and false-negative (Type II)
error rates under a critical value of 3.5, we ran two simulations. For
the false-positive error we simulated 1000 time series with 180 time
steps without introducing any anomalies and ran the \texttt{tsoutliers}
algorithm on them. Any anomalies we detect in these time series are
spurious anomalies. For the false-negative error, we simulated 1000 time
series with 180 time steps but introduced a single anomaly of the same
magnitude at the same time point in each time series. If the
\texttt{tsoutliers} algorithm fails to detect the anomaly, it would be
considered a false-negative error.

In our anomaly-less time series, we detect 208 anomalies on the 180,000
candidate time steps for a false-positive error rate of 0.116\%. In our
time series containing a single anomaly, we detect 817 anomalies out of
the 1000 candidate time series, for a false-negative error rate of
18.3\%. This provides a statistical power of 0.817. These results give
us confidence that the anomalies we detect in this paper are true
anomalies.

\begin{figure}
\centering
\includegraphics{../MainDocument/MainDocument_files/figure-latex/Figure1.png}
\caption{\textbf{``Toy'' example of anomaly detection using the exchange
rate between the Euro and the Swiss Franc.} The Swiss Franc was pegged
to the Euro at 1.2 until January 15 2015 when the peg was removed and
the currency was allowed to trade more freely. This is a very strong
anomly in the time series (\emph{t}-stat = -144.07) detectable using
statistical methods and visually. The countefactural estimate with the
outlier removed would simply have kept the currency exchange at 1.2
(\(\hat{y}\)). \label{explanationfigure}}
\end{figure}

@@

\textbf{\autoref{explanationfigure}} shows a toy example for anomaly
detection in a time series using the classic example of the exchange
rate between the Euro and the Swiss Franc (CHF). On January 15 2015, the
Swiss National Bank removed the currency peg of 1.20 francs per Euro,
exposing the Franc to the volatility of the currency market. The Franc
immediately began trading a reduced rate compared to the Euro. What
would have been the Euro/CHF exchange rate had the Swiss National Bank
\emph{not} removed the peg? A simple counter-factual estimate would be
to keep the exchange rate at 1.20 (\(\hat{y}\), sometimes called a
``synthetic control'' \citep{abadie2010synthetic}).

In the Swiss Franc example, we have knowledge of the Swiss Bank's
activities after the fact or ex-post-facto to to create the
counter-factual time series \(\hat{y}\) of 1.20. But is this anomaly
detectable without knowledge of the Swiss Bank's activities? In other
words, can we detect the reduction of the exchange rate on January 15
using only the time series? Absolutely. the tsoutliers package
identifies January 15 2015 as an extremely strong level-shift outlier
(\emph{t}-stat = -144.07). The real world is hardly this simplified
where a direct intervention is known and is testable.

@@

@@

All data and code necessary to reproduce the reported results are
licensed under the CC-BY-4.0 license and are publicly available in a
replication repository located at
\url{https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1}.
The analyses were performed in \emph{R} \citep{rcite}.

@@

We detect numerous anomalous mortality and fertility events at the US
state-level since 1999. A full listing of these anomalies can be found
in the \textbf{Supplementary Materials}. We begin by highlighting
significant fertility and mortality anomalies across all three types of
outliers (Additive Outliers, Level Shift Outliers, and Temporary Change
outliers) with plausible explanations. We then highlight two strong
anomalies that bely explanation. Finally, we conclude with a summary of
the anomalies we detect.

@@

\begin{figure}
\centering
\includegraphics{MainDocument_files/figure-latex/FertilityAnomalies-1.pdf}
\caption{\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.}
The top part of each panel contains the original time series (light
gray), the corrected, counter-factual time series in the absence of
anomalies (blue), and the red dots correspond to the onset of detected
anomalies. The bottom part of each panel contains the magnitude and type
of the outlier in red. In (a), we detect two outliers, back-to-back,
likely resulting from Hurricane Katrina in August and October 2005,
representing a decrease of more than 4,400 births due to the hurricane
(t-statistics= -6.002 and -4.997). In (b), we detect one outlier, a
level shift outlier (LS) in August 2009. We believe this reduction is
attributable to the stock market crash 11 months earlier (t-statistic =
-3.71). \label{fig:fertla}}
\end{figure}

@@

In \textbf{\autoref{fig:mortnewhamp}a}, notice the strong level shift
(LS) that occurs in February 2004 (2004:02 \emph{t}= -5.594) which
prevented 869 deaths per month. This shift totals more than 144,000
averted deaths compared to the counter-factual time series and is the
single largest mortality protective anomaly among all states. This
translates to 7\% fewer deaths than expected over the time period. What
is driving this mortality protection? What policies did NY put into
place that might have contributed to this considerable mortality
reduction? What environmental or economic conditions may have changed
may have changed? These are the kinds of questions that arise from our
analyses. The purpose of our paper is not to answer these questions, but
our findings underscore the need for more research that takes an
abductive approach. By identifying anomalies through an inductive
process, researchers can then look for underlying causes. Once those
causes are identified, researchers can then use a deductive process to
see if such events predict other (or future) anomalies.

@@

This is contrast to \textbf{\autoref{fig:ferthawaii}b}, where we
identify a strange mortality reduction in Ohio (\emph{t}-stat: -4.27).
This LS is more than 1,157 deaths per month less than the
counter-factual time series, suggesting had nearly 40,000 fewer deaths
since February 2015 than expected. This is the single-largest LS among
all states. We could not identify the potential policies Ohio might have
put into place to provide such a strong mortality protection.

@@

Consistent with more anomalies across the time series, we find more
anomalous deaths (627,364) than anomalous births (226,476). Fertility
anomalies overwhelmingly tend toward lower fertility, with only a few
fertility anomalies yielding more births. This is consistent with
research findings that link high severity weather events to
significantly lower fertility \citep{Evans2008Hurricanebirth}.
Conversely, mortality anomalies tend to be more evenly split between
positive and negative anomalies, but tend toward more deaths rather than
fewer deaths. Again, this is not surprising, as the kinds of events that
increase death (e.g., plant closings are associated with increased
opioid death \citep{Venkataramani2019}) are more common than the events
that decrease death (e.g., cannabis legalization is associated with
decreased opioid deaths \citep{Livingston2017Cannabis}). These anomalous
deaths and births account for 1.44\% and 0.889\% over the entire time
series, respectively.

As \textbf{\autoref{fig:summap}} shows, three states exhibited neither
mortality nor fertility anomalies: Alaska, North Dakota, and South
Dakota. Another nine states exhibited just a single anomaly (DE, DC, ID,
IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts exhibited
the most anomalies with nine.

@@

Data scientists frequently claim that the big data revolution is a
turning point in scientific discovery that will allow us to solve some
of the world's most pressing problems \citep{grimmer2015ppsp}. Social
scientists are skeptical of such claims, because they better understand
the complexities of the social world and know from experience that data
alone is not enough to solve social problems
\citep{bohon2018demography, grimmer2015ppsp}. Nonetheless, the increased
availability of data and (more importantly, we argue) the development of
advanced techniques for analyzing these data will enable important
discovery \citep{monroe2015no}.

>>>>

% Pandoc citation processing

@@

  \author{
        Mathew E. Hauer \\
    Department of Sociology, Florida State University\\
     and \\     Stephanie A. Bohon \\
    Department of Sociology, University of Tennessee - Knoxville\\
      }
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
The proliferation of big data, wider access to advanced computing
platforms, and the development of powerful statistical algorithms can
uncover hidden anomalies in social data, previously dismissed as noise.
Here, we combine causal inference techniques and abductive reasoning to
identify fertility and mortality anomalies on twenty years of complete
demographic data in the United States. We uncover real, ``hidden'' baby
booms/busts and mortality spikes/dips, distinguishable from regular
trend variations. We identify more than 22 and 156 fertility and
mortality anomalies, totaling more than 200k and 600k anomalous births
and deaths, respectively. Notable detectable mortality anomalies include
the September 11 2001 terrorist attack in New York and the emergence and
acceleration of the opioid epidemic in New Hampshire. Notable fertility
anomalies include the ``missing births'' in Louisiana after Hurricane
Katrina and the reduction in fertility behavior after the September 2008
stock market crash in Connecticut, amongst others. The combined causal
inference and abductive reasoning approach can be readily adapted to
find other, undiscovered social phenomena or to evaluate the efficacy of
important public policies.
\end{abstract}

@@

The search for causality is one of the primary motivations of social
scientists \citep{smith2011commentary, pearl2009causal}. Causal
inference, the identification of causal effects, attempts to uncover the
underlying mechanisms that result in changes in a phenomenon and has a
long history as an approach to inquiry in the social sciences
\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
is accomplished via a randomized control trial (RCT) or expert knowledge
applied to a natural experiment \citep{salganik2019bit}.

Unfortunately, both RCTs and natural experiments have drawbacks. RCTs
are often prohibitively expensive for many scientists, precluding their
widespread adoption in the social sciences \citep{west2008ajph}. RCTs
are also sometimes impossible to deploy in the ``real world.'' For
instance, an RCT to study country-level austerity measures would be
difficult, if not impossible, to implement. RCTs to study fertility
would be immoral. Natural experiments are a cheaper alternative but
provide less certainty about causation, fewer opportunities to replicate
work, and require expert knowledge of the natural experiment to be
properly employed \citep{pearl2018book}.

Additionally, both RCTs and natural experiments are based on traditional
hypothesis-based testing, the deductive scientific paradigm where a
question is first posed and scientists generate or find data to answer
the question. Such approaches, while extensively utilized and
time-tested, increase the likelihood that a researcher will miss
important phenomena that might go unnoticed because it falls outside the
hypothesis, or the study could suffer from extensive p-hacking
\citep{head2015extent, nuzzo2014scientific, ruggles2014big}. At the same
time, purely inductive data mining can result in false assertions or
spurious findings, especially in the hands of those who are not field
experts \citep{yanai_hypothesis_2020}. Abductive reasoning, sometimes
described as ``inferring cause from effect'' \citep{Crowder2017}, moves
the scientific process from the inductive to the deductive, and
sometimes back and forth, to reach conclusions \citep{bryant2014realm}.
Physical scientists use abductive reasoning more commonly than social
scientists. The discovery of Neptune in 1846 based on unexplained
perturbations of the orbit of Uranus \citep{popper2005logic}, the
discovery of galactic migration of planets \citep{gomes2005n}, and the
current generation of gravitational wave detectors such as the Laser
Interferometer Gravitational-Wave Observatory (LIGO)
\citep{harry2010advanced} demonstrate the power of the abductive
approach in the physical sciences.

In a data-rich environment, abductive reasoning can be more easily
applied to first find effects and then infer their cause ex-post. The
modern proliferation of data, advances in high performance (``super'')
computing, and the development of powerful statistical algorithms
provide an environment ripe for wedding causal inference with abductive
reasoning \citep{van2016data, zikopoulos2011}, with the potential for
researchers to find hidden or understudied social phenomena
\citep{bohon2018demography}. While the availability of big data and high
performance computing allows novel exploration of data through causal
inference
\citep{bohon2018demography, rcausalimpact, shiffrin2016drawing},
relatively few studies utilize causal inference techniques in the study
of demographic phenomena. However, understanding social phenomena using
these advances reveals important insights into society
\citep{angrist1989lifetime, mas2009peers} and allows us to better
monitor population trends \citep{nobles2019, torche2015hidden}. The rich
demographic data available in the United States make the potential
revelation of interesting and important demographic phenomena not only
possible, but extremely plausible -- even if identification of the
phenomena occurs without identifying the underlying cause.

In this paper, we combine a modern statistical outlier detection
algorithm \citep{chen1993joint}, abductive reasoning, and large
demographic data sets on more than 20 years of mortality and fertility
trends to identify anomalous demographic behavior. In essence, we
identify \emph{effects without knowledge of the cause}. We ask one
fundamental question regarding demographic anomalies: What are the
hidden baby booms/busts and mortality spikes/dips in the United States
over the last twenty years? We do not necessarily know the causes of
these anomalies and suggest causes for select anomalies. Even without
knowledge of the causes, simply identification of anomalies allows
scientists with more detailed knowledge of local population dynamics,
state-level policy making, or macro-economics to explain these phenomena
post-hoc. From explanations of phenomena that may have previously gone
unnoticed, social scientists may be able to better forecast populations
and provide policy solutions for impending problems such as climate
change.

@@

Scientific inquiry falls under one of three modes of logical reasoning:
deduction, induction, and abduction. Deductive reasoning starts with a
theory and makes a prediction of what the observations should be if that
theory is correct. The classic example of deductive reasoning
essentially goes (a) ``all men are mortal,'' (b) ``Socrates is a man,''
(c) ``therefore, Socrates is mortal.'' Deductive reasoning has a long
history in the sciences and most hypothesis testing in the social
sciences follows this logical reasoning. For example, John Snow used
deductive reasoning to stop the 1854 London cholera outbreak. He
hypothesized that cholera was waterborne and deduced that a bad water
pump was responsible for the outbreak.

Inductive reasoning is directionally opposite of deductive reasoning.
Rather than starting with a theory and making a prediction, induction
starts with an observation and makes a theory based on the likeliest
reason. For example, the Hispanic Mortality Paradox first appeared in
the scientific literature in the mid-1980s \citep{markides1986health} as
an observation that Hispanics in the United States had better health
outcomes than expected given their socioeconomic status. Based on this
observation, numerous theories arose to explain this paradox.

Both deductive and inductive reasoning are well established in science
but are not without their shortcomings. In both approaches, it is
impossible, or at least improbable, for the premise to be true and the
conclusion false. Furthermore, if a hypothesis is rejected under a
deductive or inductive approach, there is no alternative explanation --
there is either the presence or absence of support.

@@

Abductive reasoning and exploratory data analysis are best employed in a
data rich environment -- the very data environment social scientists and
demographers find themselves in the present moment. In a data rich
environment, a hypothesis is a hidden liability
\citep{yanai_hypothesis_2020} because it focuses our attention on only
that which we seek. To demonstrate this, \footnote{Yanai and Lercher
  \citeyearpar{yanai_hypothesis_2020} ask if focusing on a specific
  hypothesis prevents new discovery. By way of illustration, they
  created a simple dataset that if plotted would reveal an image of a
  gorilla waving. Students instructed to find support for specific
  hypotheses were less likely to discover the gorilla compared to those
  who were simply asked ``What do you conclude from the dataset?''}.
Focusing on specific hypotheses often prevents us from finding other
interesting phenomena. After all, it was the publication of the Hispanic
paradox itself that has generated more than thirty years of devoted
scholarship. At the same time, when paired with causal inference
techniques, abductive reasoning provides an empirical platform for the
discovery of potentially interesting social phenomena; it provides an
important tool for sifting large amounts of data to find interesting
results without the blinders of hypothesis testing.

@@

\begin{figure}
\centering
\includegraphics{MainDocument_files/figure-latex/ToyExample-1.pdf}
\caption{\textbf{An example of anomaly detection using Hurricanes Katrina and Rita in Orleans Parish Louisiana in 2005.}
Hurricane Katrina struck Louisiana in 2005 and we use it as a toy
example to illustrate our approach. This figure shows the annual time
series of total population in Orleans Parish between 1980 and 2019.
Between 1990 and 2005, Orleans Parish total population changed from 495k
to 494k, suggesting a possible `plateau' in the population (illustrated
with the dotted `counterfactual'). Hurricane Katrina and the widespread
population loss of more than 200k people represent a very strong anomaly
(t=-96.39) \label{explanationfigure}}
\end{figure}

@@

\textbf{\autoref{explanationfigure}} shows a toy example for anomaly
detection in a time series using the example of Hurricane Katrina in
Orleans Parish. On August 23 2005, Hurricane Katrina, a category 5
hurricane, struck southern Louisiana causing widespread damage and
destruction in Orleans Parish in particular. The displacement from the
Hurricane and the federal response were well documented
\citep{horiDisplacementDynamicsSouthern2009, fussellRecoveryMigrationCity2014}
and Census estimates suggest Orleans Parish lost more than 200,000
residents between 2005 and 2006. What would have been Orleans'
population estimate had Hurricane Katrina \emph{not} occurred? A simple
counter-factual estimate might keep population size just under 500,000
people (\(\hat{y}\)).

In the Hurricane Katrina example, we have knowledge of the impact of
Hurricane Katrina after the fact or ex-post-facto in order to create the
counter-factual time series \(\hat{y}\). But is this anomaly detectable
without knowledge of the Hurricane Katrina? In other words, can we
detect the reduction of Orleans Parish's population between 2005 and
2006 using only the time series? Absolutely. the tsoutliers package
identifies 2006 as an extremely strong level-shift outlier
(\emph{t}-stat = -96.39). The real world is hardly this simplified where
a direct intervention is known, is large, and is testable.

@@

Chen and Liu \citeyearpar{chen1993joint} report results of a simulation
exercise of varying critical value levels. They simulated various
anomaly-free time series and injected specific anomalies of varying
\(\sigma\) levels. For a critical value of 3.5 -- the level Chen and Liu
recommend to minimize both false-positive (Type I) and false-negative
(Type II) error rates -- they report Type I error rates between 0.0\%
and 0.1\% and Type II error rates or the statistical power between 40\%
and 60\% with injected anomalies equal to 4\(\sigma\). We follow their
lead by simulating time series to test for Type I and Type II errors. We
base our simulation on the 41 time series that do not contain any
anomalies (see \textbf{Results} below; 33 without a single fertility
anomaly and 8 without a single mortality anomaly).

To test for false-positive rates, we simulate 1000 time series for each
of the 41 state-components (fertility or mortality) using the fitted
ARIMA model for each state-component. This allows the simulated time
series to mimic the underlying true time series (see
\textbf{Supplementary Material} for the underlying code). Any anomalies
we detect in the 7,764,000 candidate time steps are spurious anomalies.
We detect 6,588 anomalies for a false-positive error rate of 0.085\%.

To test for false-negative rates, we use the underlying true,
sans-anomaly time series but introduce a single anomaly (magnitude equal
to 3.5\(\sigma\)) at the same time point in each series. If the
\texttt{tsoutliers} algorithm fails to detect the anomaly, it would be
considered a false-negative error. We detect 27,904 out of 41,000
candidate time series giving us a false-negative error rate of 31.9\%.
This provides a statistical power of 68.1\%.

Based on these tests we think it is clear that our probability of
committing a Type I error (or claiming an anomaly is true when in fact
is not) is very low. But, based on our false-negative error rate, it is
possible that the anomalies we do detect are \emph{conservative}. Our
statistical power is of just 68.1\% suggests a higher than normal
likelihood that true effects are going undetected. We are unlikely to
commit an error of commission (ie, including a false anomaly) but are
more likely to commit an error of omission (ie omitting true anomalies).
Thus our results should be considered conservative.

@@

All data and code necessary to reproduce the reported results are
licensed under the CC-BY-4.0 license and are publicly available as
\url{https://osf.io/yc3r4/?view_only=3ce3ca267c7047e0942940ebd08846d0}.
The analyses were performed in \emph{R} \citep{rcore}.

@@

We detect numerous anomalous mortality and fertility events at the US
state-level since 1999. A full listing of these anomalies can be found
in the \textbf{Supplementary Materials}. We begin by highlighting
significant fertility and mortality anomalies across all three types of
outliers (Additive Outliers, Level Shift Outliers, and Temporary Change
outliers) with plausible explanations. We then highlight two strong
anomalies that belie explanation. Finally, we conclude with a summary of
the anomalies we detect.

@@

\begin{figure}
\centering
\includegraphics{MainDocument_files/figure-latex/FertilityAnomalies-1.pdf}
\caption{\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.}
The top part of each panel contains the original time series (light
gray), the adjusted, counter-factual time series in the absence of
anomalies (blue), and the red dots correspond to the onset of detected
anomalies. The bottom part of each panel contains the magnitude and type
of the outlier in red. In (a), we detect two outliers, back-to-back,
likely resulting from Hurricane Katrina in August and October 2005,
representing a decrease of more than 4,400 births due to the hurricane
(t-statistics= -6.002 and -4.997). In (b), we detect one outlier, a
level shift outlier (LS) in August 2009. We believe this reduction is
attributable to the stock market crash 11 months earlier (t-statistic =
-3.71). \label{fig:fertla}}
\end{figure}

@@

In \textbf{\autoref{fig:mortnewhamp}a}, notice the strong level shift
(LS) that occurs in February 2004 (2004:02 \emph{t}= -5.594) which
prevented 869 deaths per month. This shift totals more than 144,000
averted deaths compared to the counter-factual time series and is the
single largest mortality protective anomaly among all states. This
translates to 7\% fewer deaths than expected over the time period. What
is driving this mortality protection? What policies did NY put into
place that might have contributed to this considerable mortality
reduction? What environmental or economic conditions may have changed?
These are the kinds of questions that arise from our analyses. The
purpose of our paper is not to answer these questions, but our findings
underscore the need for more research that takes an abductive approach.
By identifying anomalies through an inductive process, researchers can
then look for underlying causes. Once those causes are identified,
researchers can then use a deductive process to see if such events
predict other (or future) anomalies.

@@

This is contrast to \textbf{\autoref{fig:ferthawaii}b}, where we
identify a strange mortality reduction in Ohio (\emph{t}-stat: -4.27).
This LS is more than 1,157 deaths per month less than the
counter-factual time series, suggesting Ohio had nearly 40,000 fewer
deaths since February 2015 than expected. This is the single-largest LS
among all states. We could not identify the potential policies Ohio
might have put into place or the events that occurred to provide such a
strong mortality protection.

@@

Consistent with more anomalies across the time series, we find more
anomalous deaths (627,364) than anomalous births (226,476). Fertility
anomalies overwhelmingly tend toward lower fertility, with only a few
fertility anomalies yielding more births. This is consistent with
research findings that link severe weather events to significantly lower
fertility \citep{Evans2008Hurricanebirth}. Conversely, mortality
anomalies tend to be more evenly split between positive and negative
anomalies, but tend toward more deaths rather than fewer deaths. Again,
this is not surprising, as the kinds of events that increase death
(e.g., plant closings are associated with increased opioid death
\citep{Venkataramani2019}) are more common than the events that decrease
death (e.g., cannabis legalization is associated with decreased opioid
deaths \citep{Livingston2017Cannabis}). These anomalous deaths and
births account for 1.44\% and 0.889\% over the entire time series,
respectively.

As \textbf{\autoref{fig:summap}} shows, three states exhibited neither
mortality nor fertility anomalies: Alaska, North Dakota, and South
Dakota. Another nine states exhibited just a single anomaly (Delaware,
District of Colombia, Idaho, Illinois, Kansas, Nevada, New Mexico, Utah,
and Wyoming). Both New York and Massachusetts exhibited the most
anomalies with nine.

@@

Data scientists frequently claim that the big data revolution is a
turning point in scientific discovery that will allow us to solve some
of the world's most pressing problems \citep{grimmer2015ppsp}. Social
scientists are skeptical of such claims, because they better understand
the complexities of the social world and know from experience that data
alone are not enough to solve social problems
\citep{bohon2018demography, grimmer2015ppsp}. Nonetheless, the increased
availability of data and (more importantly, we argue) the development of
advanced techniques for analyzing these data will enable important
discovery \citep{monroe2015no}.
